Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-06-04 paper/main.tex
make corrections after meeting

\paragraph{TODO}
\begin{itemize}
    \item Change all parameter-dependent matrices to $\mtx{B}(t)$ to avoid confusion with $\mtx{A}$
    \item Change citation style to \textbf{[1]}, \textbf{[2]}, \dots
    \item Re-run numerical experiments with simplified and improved implementation (test pinvh against lstsq)
\end{itemize}
}

@@

for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This translates the problem to computing the trace of a continuous parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive. Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors. For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}. Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation}, but they have not yet been analyzed. A related technique is dynamic trace estimation for sequences $\mtx{B}_1, \dots, \mtx{B}_m$ of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}, which achieves guarantees for the estimate $b_i$ of $\Trace(\mtx{B}_i)$ of the form $\mathbb{P}(|\Trace(\mtx{B}_i) - b_i| \geq \varepsilon) \leq \delta$ for $i=1,\dots, m$. A drawback of this type of bound is that when we aim for $\varepsilon$ accuracy for all $i=1,\dots, m$ simultaneously, the probability will grow linearly with $m$ by the union bound. Therefore, instead of discretizing $[a, b]$ with $t_1, \ldots, t_m$ and looking at the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way to quantify the error is to use $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate for each $t \in [a, b]$.

@@

\paragraph{Notation}
We follow the notation and conventions which are used in contemporary literature in this field. Further, we denote with $\mathbb{N}$ all positive whole numbers; $e$ denotes Euler's number and $\log$ is the corresponding logartihm; the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$; scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots), vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots), and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots); the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$; the $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\sfrac{1}{k}}$; the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$ carries ones on its diagonal and zeros everywhere else; the zero matrix $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries; in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution; matrix norms are denoted with $\lVert \cdot \rVert$; for symmetric matrices, the $p$-Schatten norm is $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$; we refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\sfrac{1}{2}}$.

@@

We consider parameter-dependent matrices of the form
\begin{equation}
    \mtx{A}(t) = \begin{bmatrix}
        a_{11}(t) & a_{12}(t) & \dots & a_{1n}(t) \\
        a_{21}(t) & a_{22}(t) & \dots & a_{2n}(t) \\
        \vdots & \vdots & \ddots & \vdots \\
        a_{n1}(t) & a_{n2}(t) & \dots & a_{nn}(t) \\
    \end{bmatrix} \in \mathbb{R}^{n \times n}
\end{equation}
where $a_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
\begin{equation}
    \Trace(\mtx{A}(t)) = \sum_{i=1}^{n} a_{ii}(t).
    \label{equ:definition-trace}
\end{equation}
However, we assume that we only have access to products of this matrix with vectors for each $t \in [a, b]$, so this definition will not be directly useful for computing the trace.

\paragraph{Girard-Hutchinson estimator} We can estimate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
\begin{equation}
    \Hutch{\mtx{\Psi}}(\mtx{A}(t))
    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j
    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})
    \label{equ:hutchinson-trace-estimator}
\end{equation}
where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere. However, our theoretical developments only hold in the Gaussian case.

\paragraph{Nyström approximation} Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
\begin{equation}
    \Nystr{\mtx{\Omega}}{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
    \label{equ:nystrom-approximation}
\end{equation}
Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t))$. Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as

\begin{equation}
    \Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
    \label{equ:nystrom-trace-estimator}
\end{equation}

\paragraph{Nyström++ estimator} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
\begin{equation}
    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)).
    \label{equ:nystrompp-trace-estimator}
\end{equation}
This is the analogue of the Nyström++ estimator \cite{persson-2022-improved-variants} in the parameter-dependent setting. %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.

\paragraph{$L^{1}$ norm}
We define the $L^{1}$ norm of a measurable function $f:[a,b] \to \mathbb{R}$ as
\begin{equation}
    \int_{a}^{b} | f(t) | \mathrm{d}t.
\end{equation}
We use it to measure the error of the trace estimators. The following sections provide bounds on this error for each of the above-mentioned estimators.

@@

We first consider the estimator \refequ{equ:hutchinson-trace-estimator}.

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
\end{theorem}

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query Girard-Hutchinson estimate 
    \begin{equation}
        r(t) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
    \end{equation}
    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

    First, we consider $r(t)$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$. Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r^{2 k} \right]
        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments of $r$ for a fixed $t$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r \right]
        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(t)|~\mathrm{d}t  \right]
        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(t) \right]~\mathrm{d}t
        < 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(t)|~\mathrm{d}t < 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:hutchinson-trace-onequery-uniform}
    \end{equation}

    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[theorem 1]{cortinovis-2022-randomized-trace}. Let
    \begin{equation}
        \mtx{B}(t)
        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
            \mtx{A}(t) & & \\
            & \ddots & \\
            & & \mtx{A}(t)
        \end{pmatrix}
        \quad \text{and} \quad
        \vct{\psi} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\mtx{\Psi}}}
        \end{pmatrix}
    \end{equation}
    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors. Then $\lVert \mtx{B}(t) \rVert _F = \lVert \mtx{A}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ and
    \begin{equation}
        r(t) = \Trace(\mtx{A}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
    \end{equation}
    Hence, we conclude from \refequ{equ:hutchinson-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
        < 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
    \end{equation}

    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
    \begin{equation}
        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log(\delta^{-1}) \rceil}}
        = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
        \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
    \end{equation}
    from which follows the second part of the theorem.
    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
    %\end{equation}
\end{proof}

>>>>

\paragraph{TODO}
\begin{itemize}
    \item Change citation style to \textbf{[1]}, \textbf{[2]}, \dots
    \item Re-run numerical experiments with simplified and improved implementation (test pinvh against lstsq)
\end{itemize}
}

@@

for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This translates the problem to computing the trace of a continuous parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive. Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors. For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}. Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation}, but they have not yet been analyzed. \todo{A related technique is dynamic trace estimation for sequences $\mtx{B}_1, \dots, \mtx{B}_m$ of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}, which achieves guarantees for the estimate $b_i$ of $\Trace(\mtx{B}_i)$ of the form $\mathbb{P}(|\Trace(\mtx{B}_i) - b_i| \geq \varepsilon) \leq \delta$ for $i=1,\dots, m$. A drawback of this type of bound is that when we aim for $\varepsilon$ accuracy for all $i=1,\dots, m$ simultaneously, the probability will grow linearly with $m$ by the union bound. Therefore, instead of discretizing $[a, b]$ with $t_1, \ldots, t_m$ and looking at the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way to quantify the error is to use $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate for each $t \in [a, b]$.}

@@

@@

We consider parameter-dependent matrices of the form
\begin{equation}
    \mtx{B}(t) = \begin{bmatrix}
        b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
        b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
        \vdots & \vdots & \ddots & \vdots \\
        b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
    \end{bmatrix} \in \mathbb{R}^{n \times n}
\end{equation}
where $b_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
\begin{equation}
    \Trace(\mtx{B}(t)) = \sum_{i=1}^{n} b_{ii}(t).
    \label{equ:definition-trace}
\end{equation}
However, we assume that we only have access to products of this matrix with vectors for each $t \in [a, b]$, so this definition will not be directly useful for computing the trace.

\paragraph{Girard-Hutchinson estimator} We can estimate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
\begin{equation}
    \Hutch{\mtx{\Psi}}(\mtx{B}(t))
    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi})
    \label{equ:hutchinson-trace-estimator}
\end{equation}
where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere. However, our theoretical developments only hold in the Gaussian case.

\paragraph{Nyström approximation} Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
\begin{equation}
    \Nystr{\mtx{\Omega}}{\mtx{B}}(t) = (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
    \label{equ:nystrom-approximation}
\end{equation}
Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t))$. Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as

\begin{equation}
    \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
    \label{equ:nystrom-trace-estimator}
\end{equation}

\paragraph{Nyström++ estimator} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
\begin{equation}
    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)).
    \label{equ:nystrompp-trace-estimator}
\end{equation}
This is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}. %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.

@@

We first consider the estimator \refequ{equ:hutchinson-trace-estimator}. \todo{Introduce $L^1$ norm and Frobenius norm}

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t < 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
\end{theorem}

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query Girard-Hutchinson estimate 
    \begin{equation}
        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
    \end{equation}
    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ \todo{introduce spectral norm}. Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
        < 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t < 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:hutchinson-trace-onequery-uniform}
    \end{equation}

    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[theorem 1]{cortinovis-2022-randomized-trace}. Let
    \begin{equation}
        \widetilde{\mtx{B}}(t)
        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
            \mtx{B}(t) & & \\
            & \ddots & \\
            & & \mtx{B}(t)
        \end{pmatrix}
        \quad \text{and} \quad
        \widetilde{\vct{\psi}} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\mtx{\Psi}}}
        \end{pmatrix}
    \end{equation}
    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors. Then $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$, and
    \begin{equation}
        r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
    \end{equation}
    Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
        < 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
